An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets

### 2022-04-12 rj_format/rj-format-ju-hofmann/ju-hofmann.tex
change to rjtools

\newcommand{\hh}[1]{{\textcolor{orange}{#1}}}
\newcommand{\wj}[1]{{\textcolor{blue}{#1}}}
\setlength{\parindent}{0pt}

@@

In this paper we present an open-source implementation of the algorithm
of the Congruent Matching Profile Segments (CMPS) method. \citet{cmps}
developed the CMPS method for ``objective comparison of striated tool
marks'' and demonstrated its use in some examples of comparing bullet
signature correlations. Although \citet{cmps} conceptually describe the
CMPS algorithm in their paper, the authors did not release an
implementation of their method. Thus, our effort here is to introduce
the CMPS method to the \texttt{R} \citep{R} community and provide an
open-source, publicly available implementation of the algorithm to use,
review, and improve. Our implementation is made available as part of the
\texttt{R} package \pkg{cmpsR} on CRAN.

According to the Uniform Crime Reporting Program of FBI (Federal Bureau
of Investigation), ``more than 76 percent (76.7) of the homicides for
which the FBI received weapons data in 2020 involved the use of
firearms'' \citep{fbiucr}. At the same time, the number of murder
victims jumped between 2019 and 2020 by more than 23 percent (23.4) to
17754 \citep{fbiucr20}. This increase is unprecedented and highlights
the important role that firearm examination plays in all these cases. An
important task of firearm examination is to answer the question of
whether two pieces of evidence come from the same source or whether a
piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets
were fired from the same gun barrel. Assessing the similarity between
two bullets is based on a comparison of striation marks acquired during
the firing process as bullets are propelled through the barrel. Current
state of art sees firearms examiners make an assessment of similarity
based on a visual comparison, generally, using a comparison microscope
\citep{afte}. This practice has been criticized for its lack of
objectivity and the associated problem of determining valid error rates
\citep{pcast}.

A report published by the \citet{nrc} states that ``{[}m{]}uch forensic
evidence-including, for example, bite marks and firearm and toolmark
identification---is introduced in criminal trials without any meaningful
scientific validation, determination of error rates, or reliability
testing to explain the limits of the discipline.'' To overcome those
criticisms and concerns, researchers have been making an effort to build
databases and develop frameworks and algorithms that bring an objective
and quantitative assessment into the field. The database of
\citet{nistdb} with digital 3D topographic scans from various studies by
\citet{brundage}, \citet{hamby}, \citet{Hamby:2019} and others, provides
a great resource for researchers. Algorithms that can be validated and
tested and generate quantitative results are developed for 2D and 3D
surface texture \citep{song2005}, tool marks
\citep{ChumbleyL_Scott2010VoTM}, and striae on Land Engraved Areas
(LEAs) of bullets \citep[\citet{pmid30444940}, \citet{cmps}]{aoas}. The
notion of bullet signatures is one of the results of these efforts, and
we will use it to explain the CMPS algorithm and showcase the
\pkg{cmpsR} implementation. Bullet signatures can be extracted from
bullet LEAs, typically one for each land engraved area, and how those
bullet signatures are extracted from scans will be discussed in the
Background section. Bullet signatures capture striation marks on the
bullet in a numeric format and therefore serve as the foundation for
algorithms. These scans played an important role in bringing objectivity
into the field. In the following sections we will: review the background
of bullet signature comparisons, discuss how we followed the idea
described by \citet{cmps} for the implementation of the CMPS algorithm,
propose new metrics based on the CMPS score and a principled evaluation
framework for algorithmic results comparison, and present results of
applying our implementation to real data.

@@

The datasets we worked with come from the James Hamby Consecutively
Rifled Ruger Barrel Study \citep{brundage, hamby, Hamby:2019}, in
particular, Hamby set 252 and Hamby set 44. For each Hamby set, a total
of 35 bullets is fired from (the same) ten consecutively manufactured
Ruger P-85 pistol barrels. Two bullets are fired from each barrel,
making up a set of 20 reference bullets. An additional 15 bullets are
fired from these ten barrels in a fashion unknown to the study
participant. The aim of the Hamby Study was to have firearms examiners
identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves
and lands as shown in \autoref{fig:bullet}. During the firing process,
grooves and lands are engraved on a bullet. Firearms examiners use
striation marks on land engraved areas (LEAs) for their visual
comparison. For algorithmic purposes, 3D topographical images of land
engraved areas were obtained and stored in x3p format (XML 3-D Surface
Profile). The x3p format provides a standard way of exchanging 2D and 3D
profile data. It conforms to
\href{http://sourceforge.net/p/open-gps/mwiki/X3p/}{the ISO5436-2
standard} adopted by the OpenFMC (Open Forensic Metrology Consortium), a
group of firearm forensics researchers who contributes to the
establishment of best practices of using metrology in forensic science.
Hamby set 252 was scanned using a NanoFocus lens at 20x magnification
with the scan resolution being 1.5625 \textmu m \(\times\) 1.5625
\textmu m per pixel. Hamby set 44 was scanned at the Roy J Carver
High-Resolution microscopy lab at Iowa State. These scans were acquired
with a Sensofar Confocal Light Microscope at 20x magnification for a
nominal resolution of 0.645 \textmu m \(\times\) 0.645 \textmu m per
pixel. Both Hamby set 252 and Hamby set 44 are publicly available from
the NIST Ballistics Database Project \citep{nistdb}.

@@

The automated framework for extracting signatures from x3p files used in
this paper was proposed by \citet{aoas} and is implemented in the
\texttt{R} packages \pkg{x3ptools} \citep{x3ptools} and
\pkg{bulletxtrctr} \citep{bulletxtrctr}. \pkg{x3ptools} is a package to
read, write, and generally, process x3p files. The \pkg{bulletxtrctr}
package implements a pipeline for extracting and comparing signatures
from scans of land-engraved areas.

\autoref{fig:process} gives an overview of all of the steps in the
process from scan to signatures. \autoref{fig:process}(a) shows a
rendering of a 3D scan of a bullet land engraved area. The raised
portion of the surface on the left and right of the scan are parts of
the adjacent groove engraved areas (GEAs), the middle area shows a
land-engraved area (LEA) with well expressed striation marks. The first
step of obtaining the bullet signature is to extract a cross-section at
a fixed height along the land engraving.

The thin white horizontal line in \autoref{fig:process}(a) indicates
which cross-section was identified by the algorithm to represent the
LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional
view. Groove engraved areas are removed from the analysis as indicated
by the vertical blue lines in \autoref{fig:process}(c) and
\autoref{fig:process}(d)). A non-parametric LOESS smooth \citep{loess}
is fitted to capture the bullet curvature (\autoref{fig:process}(e))
and, finally, the \textbf{bullet signature} (\autoref{fig:process}(f))
is obtained as residuals of the cross-section and the fitted smooth.
Note, that in \citet{cmps} bullet signatures are referred to as bullet
profiles. However, to avoid confusion, we distinguish the notion of
bullet signatures from bullet profiles. Bullet profiles are shown in
panels (b), (c), and (d) of \autoref{fig:process}, while
\autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a
correct down-stream analysis of the signatures. To allow for a
human-in-the middle inspection and intervention we have provided an
interactive web-application, implemented as \texttt{R} Shiny App
\citep{shiny}, \pkg{bulletinspectR} to identify and correct those
errors. An example of the extraction process with corresponding code and
parameter settings can be found in the ``Supplementary materials''. Note
that the process of extracting signatures might be different from the
one used in \citet{cmps} because no code or parameter settings are made
available publicly.

@@

\caption[A framework of obtaining a bullet signature]{A framework of obtaining a bullet signature. (a) rendering from the 3D topographic scan of a land engraved area (LEA). The selected crosscut location is indicated by a thin white horizontal line. (b) view of the cross-section of the land engraved area at the white line in (a). (c) the crosscut data plotted in 2D; blue vertical lines indicate the position of left and right grooves. (d) the crosscut data after chopping the left and right grooves. (e) the fitted curvature using LOESS. (f) after removing the curvature from the crosscut data, the bullet signature is obtained}\label{fig:process}
\end{figure}
\end{Schunk}

@@

In this example, the comparison between \code{sig1} and \code{sig2}, two
signatures coming from the same source (a known-match comparison), gets
a CMPS score of 17; the comparison between \code{sig1} and \code{sig3},
two signatures coming from different sources (a known non-match
comparison), gets a CMPS score of 1.

@@

The section ``Implementation'' will go through the algorithm and show
how to use the \pkg{cmpsR} package. A further example that illustrates
the main points is also included. In the section on 'Evaluation
metrics'' we introduce new CMPS metrics that summarize land-level CMPS
scores and a sum of squares ratio that can be used to evaluate
algorithmic results. The section ``Results'' presents the results of
evaluating the \pkg{cmpsR} package using Hamby set 252 and Hamby set 44.
The results from Hamby set 252 are used to verify that our
implementation is, at least qualitatively, comparable to the algorithm
described in \citet{cmps}. Results from Hamby 44 show the need for a
further investigation of the parameter choices even in the case of
bullets fired from the same barrels. The last section covers some final
discussion and conclusions.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{cut the comparison signature into consecutive, non-overlapping
  and equal-length basis segments:} The command
  \code{get\_segs(x, len=50)} implements this step: it takes bullet
  signature \code{x} in the format of a numeric vector and cuts it into
  consecutive, non-overlapping and equal-length segments of length
  \code{len}, which are referred to as ``basis segments''. Note that the
  parameter \code{len} determines the length of a basis segment and thus
  affects the total number of basis segments, which is the upper limit
  of the CMPS score of a comparison. The default value of \code{len} is
  50, which will result in about 25 basis segments for the example data
  of the package.
\item
  \textbf{identify candidate positions:} For each basis segment a set of
  candidate registration positions on the comparison signature is
  identified based on the segment's similarity to the reference
  signature. In a first step, the cross-correlation function of the
  segment to the reference is calculated, then a number of positions
  with high correlation values are identified as candidate positions. In
  case multiple segment lengths are considered, the length of each basis
  segment is expanded (by default it is doubled) and these two steps are
  repeated. Only when candidate positions coincide (or are similar
  enough), they are considered further. \autoref{fig:segplots} and
  \autoref{fig:seg_all} illustrate these ideas.

  \begin{itemize}
  \item
    \textbf{Calculate the cross-correlation curve:} Calculate the
    cross-correlation curve between a basis segment \code{x} and the
    reference signature \code{y} using the function
    \code{get\_ccf4(x, y, ...)} as shown in \autoref{fig:segplot2}. The
    position indicates the lag by which a basis segment is moved with
    respect to its original placement. A position is considered ``good''
    if it results in a peak in the cross-correlation between the basis
    segment and the reference.
  \item
    \textbf{Correlation peaks:} Two strategies referred to as
    ``multi-peak inspection'' and ``multi-peak inspection at different
    segment lengths'' in \citet{cmps} are used for identifying positions
    of correlation peaks as candidate positions. The latter is also
    called ``multi-segment lengths strategy''. The parameter
    \code{npeaks\_set} in \code{extract\_feature\_cmps(...)} determines
    which strategy to use and the number of candidate positions:

    \begin{itemize}
    \tightlist
    \item
      If \code{npeaks\_set} is an integer vector of length 1, for
      example \code{npeaks\_set = 5}, the positions of the top five
      peaks in the cross-correlation curve are identified as candidate
      positions for registration.
    \item
      If \code{npeaks\_set} is an integer vector of length more than 1,
      for example \code{npeaks\_set = c(5, 3, 1)}, the multi-segment
      lengths strategy will be used: calculate the cross-correlation
      function between a basis segment and the reference and identify
      positions of the top five peaks; double the segment length to a
      specified value, re-calculate the cross-correlation function, and
      identify three peaks; repeat this process and identify a single
      peak in the newly computed cross-correlation function.
      \autoref{fig:seg_all} shows an example of three levels of basis
      segment 6 and their corresponding cross-correlation curves and
      identified peaks. Note that in \citet{cmps} the segment length is
      doubled at each level of a basis segment, but in the present
      implementation users are allowed to choose the segment length at
      each level.
    \item
      \code{get\_ccr\_peaks(comp, segments, seg\_outlength, nseg = 1, npeaks = 5)}
      computes the cross-correlation curve between a basis or increased
      segment and the reference signature and finds peaks in the
      cross-correlation curve. The number of peaks detected is equal to
      \code{npeaks}, which is an integer. \code{segments},
      \code{seg\_outlength}, and \code{nseg} determines the segment in
      the cross-correlation computation, and \code{comp} gives the
      reference signature. If the multi-segment lengths strategy is
      used, then \code{get\_ccr\_peaks(...)} is called in a
      \code{lapply()} for each level of the basis segment. The resulting
      list is called \code{ccr\_list}.
    \end{itemize}
  \item
    \textbf{multi-segment lengths strategy:} with the multi-segment
    lengths strategy being used, a position is identified as a candidate
    position for registration and is called as a ``consistent
    correlation peak'' if it results in a top peak in the
    cross-correlation curve with a tolerance zone determined by
    \code{Tx} in all segment levels. Note that in \citet{cmps}, a
    segment at its largest scale (highest level) always identifies one
    peak, but we do not have this requirement in our implementation.

    \begin{itemize}
    \tightlist
    \item
      the function \code{get\_seg\_scale(segments, nseg, out\_length)}
      is used to obtain the (potentially increased) version of a basis
      segment. \code{segments}, which is a list containing all basis
      segments generated by the function \code{get\_segs(...)} in step
      1, and \code{nseg} are used to determine the basis segment to be
      increased. \code{out\_length} specifies the length of the output
      segment.
    \item
      \code{get\_ccp(ccr\_list, Tx = 25)} tries to identify the
      ``consistent correlation peak''. \code{ccr\_list} is the result of
      \code{lapply()} and \code{get\_ccr\_peaks(...)}, and \code{Tx}
      determines the size of a tolerance zone used in identifying the
      consistent correlation peak. \code{get\_ccp(...)} returns
      \code{NULL} if there is no consistent correlation peak.
    \end{itemize}
  \end{itemize}
\item
  \textbf{determine the congruent registration position:} A candidate
  position ``receives'' votes from basis segments that identify it or a
  close position within a tolerance zone of \code{Tx} as a candidate
  position in step 2. Votes for all candidate positions are tallied, and
  the position with the highest number of votes gets chosen as the
  \emph{congruent registration position}, indicating that most of the
  basis segments find their highly similar counterpart in the reference
  signature in terms of correlation at this registration position. In
  case of ties the middle position is taken as the congruent
  registration position. Basis segments with a congruent registration
  position are called ``congruent matching profile segments'' (CMPS).
  The total number of CMPS is the CMPS score of the comparison.
  \code{get\_CMPS(input\_ccp, Tx = 25)} is the function that tallies the
  votes and determines the congruent registration position and congruent
  matching profile segments (CMPS).
\end{enumerate}

Note that there are several parameters in the CMPS algorithm that will
affect the final results and are left to the users to decide, such as
the length of basis segments \code{seg\_length} in step 1, the number of
peaks \code{npeaks\_set} identified on each level in step 2, and the
length of the tolerance zone \code{Tx} in both step 2 and 3. In our
implementation of the CMPS algorithm we used the parameters given in the
original CMPS paper \citep{cmps} as the default values for these
parameters. However, the authors state that no cross-validation has been
done - there might also be issues with respect to resolution of the
scans. Further research is needed, until then users are advised to think
of default values as starting values and consider alternatives. However,
the evaluation framework based on the Sum of Squares Ratio introduced in
the later section could be used to evaluate the choices of these
parameters. The main function that combines all steps in the CMPS
algorithm described above is called \code{extract\_feature\_cmps(...)}.
Here we present it with its default parameters.

@@

The function \code{extract\_feature\_cmps} allows for the following
input from users besides the previously discussed parameters
\code{seg\_length}, \code{npeaks\_set}, and \code{Tx}:

\begin{itemize}
\item
  \code{x} and \code{y} are two signatures: \code{x} serves as the
  comparison signature (which will be divided into basis segments) and
  \code{y} is the reference signature;
\item
  \code{include} determines the format of the function result. Besides
  the \code{CMPS\_score}, other aspects of the comparison help in
  understanding how the \code{CMPS\_score} is computed. By default
  \code{include} is set to \code{NULL} and only the \code{CMPS\_score}
  is returned; further results are included when \code{include} is (an
  abbreviation of) one of or a vector of the following strings:
  \code{"nseg"}, \code{"congruent\_pos"}, \code{"congruent\_seg\_idx"},
  \code{"segments"}, \code{"parameters"}, and \code{"full\_result"}. If
  \code{include} is specified as \code{"full\_result"} (or its
  abbreviation), the output includes everything listed below.

  \begin{itemize}
  \tightlist
  \item
    \code{nseg}: the number of basis segments from the comparison
    signature; this is also the highest possible CMPS score of the
    comparison;
  \item
    \code{congruent\_pos}: the congruent registration position;
  \item
    \code{congruent\_seg\_idx}: the indices of all congruent matching
    profile segments;
  \item
    \code{ccp\_list}: a list showing identified candidate positions of
    all basis segments;
  \item
    \code{pos\_df}: a data frame containing all candidate positions and
    their respective number of votes;
  \item
    \code{segments}: a list containing all basis segments;
  \item
    \code{parameters}: a list containing all input arguments of
    \code{extract\_feature\_cmps};
  \end{itemize}
\item
  \code{outlength} specifies the segment length of a basis segment at
  each level under the multi-segment lengths strategy. By default
  \code{outlength} is set to \code{NULL}, indicating that a basis
  segment should double its segment length at the next level and
  conforming to the description in \citet{cmps} .
\end{itemize}

In the remainder of the paper we showcase the use of the \pkg{cmpsR}
functionality on some examples and present the results of applying it to
two datasets.

>>>>

@@

In this paper, we present an open-source implementation of the algorithm
of the Congruent Matching Profile Segments (CMPS) method. \citet{cmps}
developed the CMPS method for ``objective comparison of striated tool
marks'' and demonstrated its use in some examples of comparing bullet
signature correlations. Although \citet{cmps} conceptually describe the
CMPS algorithm in their paper, the authors did not release an
implementation of their method. Thus, our effort here is to introduce
the CMPS method to the \texttt{R} \citep{R} community and provide an
open-source, publicly available implementation of the algorithm to use,
review, and improve. Our implementation is made available as part of the
\texttt{R} package \CRANpkg{cmpsR} on CRAN.

According to the Uniform Crime Reporting Program of the FBI (Federal
Bureau of Investigation), ``more than 76 percent (76.7) of the homicides
for which the FBI received weapons data in 2020 involved the use of
firearms'' \citep{fbiucr}. At the same time, the number of murder
victims jumped between 2019 and 2020 by more than 23 percent (23.4) to
17754 \citep{fbiucr20}. This increase is unprecedented and highlights
the important role that firearm examination plays in all these cases. An
important task of firearm examination is to answer the question of
whether two pieces of evidence come from the same source or whether a
piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets
were fired from the same gun barrel. Assessing the similarity between
two bullets is based on a comparison of striation marks acquired during
the firing process as bullets are propelled through the barrel. The
current state of art sees firearms examiners make an assessment of
similarity based on a visual comparison, generally, using a comparison
microscope \citep{afte}. This practice has been criticized for its lack
of objectivity and the associated problem of determining valid error
rates \citep{pcast}.

A report published by the \citet{nrc} states that ``{[}m{]}uch forensic
evidence-including, for example, bite marks and firearm and toolmark
identification---is introduced in criminal trials without any meaningful
scientific validation, determination of error rates, or reliability
testing to explain the limits of the discipline.'' To overcome those
criticisms and concerns, researchers have been making an effort to build
databases and develop frameworks and algorithms that bring an objective
and quantitative assessment into the field. The database of
\citet{nistdb} with digital 3D topographic scans from various studies by
\citet{brundage}, \citet{hamby}, \citet{Hamby:2019} and others, provides
a great resource for researchers. Algorithms that can be validated and
tested and generate quantitative results are developed for 2D and 3D
surface texture \citep{song2005}, tool marks
\citep{ChumbleyL_Scott2010VoTM}, and striae on Land Engraved Areas
(LEAs) of bullets \citep[\citet{pmid30444940}, \citet{cmps}]{aoas}. The
notion of bullet signatures is one of the results of these efforts, and
we will use it to explain the CMPS algorithm and showcase the
\pkg{cmpsR} implementation. Bullet signatures can be extracted from
bullet LEAs, typically one for each land engraved area, and how those
bullet signatures are extracted from scans will be discussed in the
Background section. Bullet signatures capture striation marks on the
bullet in a numeric format and therefore serve as the foundation for
algorithms. These scans played an important role in bringing objectivity
into the field. In the following sections, we will: review the
background of bullet signature comparisons, discuss how we followed the
idea described by \citet{cmps} for the implementation of the CMPS
algorithm, propose new metrics based on the CMPS score and a principled
evaluation framework for algorithmic results comparison, and present
results of applying our implementation to real data.

@@

The datasets we worked with come from the James Hamby Consecutively
Rifled Ruger Barrel Study \citep{brundage, hamby, Hamby:2019}, in
particular, Hamby set 252 and Hamby set 44. For each Hamby set, a total
of 35 bullets is fired from (the same) ten consecutively manufactured
Ruger P-85 pistol barrels. Two bullets are fired from each barrel,
making up a set of 20 reference bullets. An additional 15 bullets are
fired from these ten barrels in a fashion unknown to the study
participant. The aim of the Hamby Study was to have firearms examiners
identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves
and lands as shown in Figure \ref{fig:bullet}. During the firing
process, grooves and lands are engraved on a bullet. Firearms examiners
use striation marks on land engraved areas (LEAs) for their visual
comparison. For algorithmic purposes, 3D topographical images of land
engraved areas were obtained and stored in x3p format (XML 3-D Surface
Profile). The x3p format provides a standard way of exchanging 2D and 3D
profile data. It conforms to
\href{http://sourceforge.net/p/open-gps/mwiki/X3p/}{the ISO5436-2
standard} adopted by the OpenFMC (Open Forensic Metrology Consortium), a
group of firearm forensics researchers who contributes to the
establishment of best practices of using metrology in forensic science.
Hamby set 252 was scanned using a NanoFocus lens at 20x magnification
with the scan resolution being 1.5625 \textmu m \(\times\) 1.5625
\textmu m per pixel. Hamby set 44 was scanned at the Roy J Carver
High-Resolution microscopy lab at Iowa State. These scans were acquired
with a Sensofar Confocal Light Microscope at 20x magnification for a
nominal resolution of 0.645 \textmu m \(\times\) 0.645 \textmu m per
pixel. Both Hamby set 252 and Hamby set 44 are publicly available from
the NIST Ballistics Database Project \citep{nistdb}.

@@

The automated framework for extracting signatures from x3p files used in
this paper was proposed by \citet{aoas} and is implemented in the
\texttt{R} packages \CRANpkg{x3ptools} \citep{x3ptools} and
\pkg{bulletxtrctr} \citep{bulletxtrctr}. \pkg{x3ptools} is a package to
read, write, and generally, process x3p files. The \pkg{bulletxtrctr}
package implements a pipeline for extracting and comparing signatures
from scans of land-engraved areas.

Figure \ref{fig:process} gives an overview of all of the steps in the
process from scan to signatures. Figure \ref{fig:process}(a) shows a
rendering of a 3D scan of a bullet land engraved area. The raised
portion of the surface on the left and right of the scan are parts of
the adjacent groove engraved areas (GEAs), the middle area shows a
land-engraved area (LEA) with well expressed striation marks. The first
step of obtaining the bullet signature is to extract a cross-section at
a fixed height on the land engraving.

The thin white horizontal line in Figure \ref{fig:process}(a) indicates
which cross-section was identified by the algorithm to represent the
LEA; Figure \ref{fig:process}(b) shows the corresponding cross-sectional
view. Groove engraved areas are removed from the analysis as indicated
by the vertical blue lines in Figure \ref{fig:process}(c) and Figure
\ref{fig:process}(d)). A non-parametric LOESS smooth \citep{loess} is
fitted to capture the bullet curvature ( Figure \ref{fig:process}(e))
and, finally, the \textbf{bullet signature} (Figure
\ref{fig:process}(f)) is obtained as residuals of the cross-section and
the fitted smooth. Note, that in \citet{cmps} bullet signatures are
referred to as bullet profiles. However, to avoid confusion, we
distinguish the notion of bullet signatures from bullet profiles. Bullet
profiles are shown in panels (b), (c), and (d) of Figure
\ref{fig:process}, while Figure \ref{fig:process}(f) shows the
corresponding bullet signature. Identifying the groove engraved areas
correctly is fundamental for a correct downstream analysis of the
signatures. To allow for a human-in-the-middle inspection and
intervention we have provided an interactive web application,
implemented as \texttt{R} Shiny App \citep{shiny}, \pkg{bulletinspectR}
to identify and correct those errors. An example of the extraction
process with corresponding code and parameter settings can be found in
the ``Supplementary materials''. Note that the process of extracting
signatures might be different from the one used in \citet{cmps} because
no code or parameter settings are made available publicly.

@@

\caption[A framework of obtaining a bullet signature]{A framework of obtaining a bullet signature. (a) rendering from the 3D topographic scan of a land engraved area (LEA). The selected crosscut location is indicated by a thin white horizontal line. (b) view of the cross-section of the land engraved area at the white line in (a). (c) the crosscut data plotted in 2D; blue vertical lines indicate the position of the left and right grooves. (d) the crosscut data after chopping the left and right grooves. (e) the fitted curvature using LOESS. (f) after removing the curvature from the crosscut data, the bullet signature is obtained.}\label{fig:process}
\end{figure}
\end{Schunk}

@@

In this example, the comparison between \texttt{sig1} and \texttt{sig2},
two signatures coming from the same source (a known-match comparison),
gets a CMPS score of 17; the comparison between \texttt{sig1} and
\texttt{sig3}, two signatures coming from different sources (a known
non-match comparison), gets a CMPS score of 1.

@@

The section ``Implementation'' will go through the algorithm and show
how to use the \pkg{cmpsR} package. A further example that illustrates
the main points is also included. In the section on 'Evaluation
metrics'', we introduce new CMPS metrics that summarize land-level CMPS
scores and a sum of squares ratio that can be used to evaluate
algorithmic results. The section ``Results'' presents the results of
evaluating the \pkg{cmpsR} package using Hamby set 252 and Hamby set 44.
The results from Hamby set 252 are used to verify that our
implementation is, at least qualitatively, comparable to the algorithm
described in \citet{cmps}. Results from Hamby 44 show the need for a
further investigation of the parameter choices even in the case of
bullets fired from the same barrels. The last section covers some final
discussion and conclusions.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{cut the comparison signature into consecutive,
  non-overlapping, and equal-length basis segments:} The command
  \texttt{get\_segs(x,\ len=50)} implements this step: it takes bullet
  signature \texttt{x} in the format of a numeric vector and cuts it
  into consecutive, non-overlapping and equal-length segments of length
  \texttt{len}, which are referred to as ``basis segments''. Note that
  the parameter \texttt{len} determines the length of a basis segment
  and thus affects the total number of basis segments, which is the
  upper limit of the CMPS score of a comparison. The default value of
  \texttt{len} is 50, which will result in about 25 basis segments for
  the example data of the package.
\item
  \textbf{identify candidate positions:} For each basis segment a set of
  candidate registration positions on the comparison signature is
  identified based on the segment's similarity to the reference
  signature. In the first step, the cross-correlation function of the
  segment to the reference is calculated, then a number of positions
  with high correlation values are identified as candidate positions. In
  case multiple segment lengths are considered, the length of each basis
  segment is expanded (by default it is doubled) and these two steps are
  repeated. Only when candidate positions coincide (or are similar
  enough), they are considered further. Figure \ref{fig:segplots} and
  Figure \ref{fig:segall} illustrate these ideas.

  \begin{itemize}
  \item
    \textbf{Calculate the cross-correlation curve:} Calculate the
    cross-correlation curve between a basis segment \texttt{x} and the
    reference signature \texttt{y} using the function
    \texttt{get\_ccf4(x,\ y,\ ...)} as shown in Figure
    \ref{fig:segplot2}. The position indicates the lag by which a basis
    segment is moved with respect to its original placement. A position
    is considered ``good'' if it results in a peak in the
    cross-correlation between the basis segment and the reference.
  \item
    \textbf{Correlation peaks:} Two strategies referred to as
    ``multi-peak inspection'' and ``multi-peak inspection at different
    segment lengths'' in \citet{cmps} are used for identifying positions
    of correlation peaks as candidate positions. The latter is also
    called the ``multi-segment lengths strategy''. The parameter
    \texttt{npeaks\_set} in \texttt{extract\_feature\_cmps(...)}
    determines which strategy to use and the number of candidate
    positions:

    \begin{itemize}
    \tightlist
    \item
      If \texttt{npeaks\_set} is an integer vector of length 1, for
      example, \texttt{npeaks\_set\ =\ 5}, the positions of the top five
      peaks in the cross-correlation curve are identified as candidate
      positions for registration.
    \item
      If \texttt{npeaks\_set} is an integer vector of length more than
      1, for example, \texttt{npeaks\_set\ =\ c(5,\ 3,\ 1)}, the
      multi-segment lengths strategy will be used: calculate the
      cross-correlation function between a basis segment and the
      reference and identify positions of the top five peaks; double the
      segment length to a specified value, re-calculate the
      cross-correlation function, and identify three peaks; repeat this
      process and identify a single peak in the newly computed
      cross-correlation function. Figure \ref{fig:segall} shows an
      example of three levels of basis segment 6 and their corresponding
      cross-correlation curves and identified peaks. Note that in
      \citet{cmps} the segment length is doubled at each level of a
      basis segment, but in the present implementation users are allowed
      to choose the segment length at each level.
    \item
      \texttt{get\_ccr\_peaks(comp,\ segments,\ seg\_outlength,\ nseg\ =\ 1,\ npeaks\ =\ 5)}
      computes the cross-correlation curve between a basis or increased
      segment and the reference signature and finds peaks in the
      cross-correlation curve. The number of peaks detected is equal to
      \texttt{npeaks}, which is an integer. \texttt{segments},
      \texttt{seg\_outlength}, and \texttt{nseg} determine the segment
      in the cross-correlation computation, and \texttt{comp} gives the
      reference signature. If the multi-segment lengths strategy is
      used, then \texttt{get\_ccr\_peaks(...)} is called in a
      \texttt{lapply()} for each level of the basis segment. The
      resulting list is called \texttt{ccr\_list}.
    \end{itemize}
  \item
    \textbf{multi-segment lengths strategy:} with the multi-segment
    lengths strategy being used, a position is identified as a candidate
    position for registration and is called a ``consistent correlation
    peak'' if it results in a top peak in the cross-correlation curve
    with a tolerance zone determined by \texttt{Tx} in all segment
    levels. Note that in \citet{cmps}, a segment at its largest scale
    (highest level) always identifies one peak, but we do not have this
    requirement in our implementation.

    \begin{itemize}
    \tightlist
    \item
      the function
      \texttt{get\_seg\_scale(segments,\ nseg,\ out\_length)} is used to
      obtain the (potentially increased) version of a basis segment.
      \texttt{segments}, which is a list containing all basis segments
      generated by the function \texttt{get\_segs(...)} in step 1, and
      \texttt{nseg} are used to determine the basis segment to be
      increased. \texttt{out\_length} specifies the length of the output
      segment.
    \item
      \texttt{get\_ccp(ccr\_list,\ Tx\ =\ 25)} tries to identify the
      ``consistent correlation peak''. \texttt{ccr\_list} is the result
      of \texttt{lapply()} and \texttt{get\_ccr\_peaks(...)}, and
      \texttt{Tx} determines the size of a tolerance zone used in
      identifying the consistent correlation peak.
      \texttt{get\_ccp(...)} returns \texttt{NULL} if there is no
      consistent correlation peak.
    \end{itemize}
  \end{itemize}
\item
  \textbf{determine the congruent registration position:} A candidate
  position ``receives'' votes from basis segments that identify it or a
  close position within a tolerance zone of \texttt{Tx} as a candidate
  position in step 2. Votes for all candidate positions are tallied, and
  the position with the highest number of votes gets chosen as the
  \emph{congruent registration position}, indicating that most of the
  basis segments find their highly similar counterpart in the reference
  signature in terms of correlation at this registration position. In
  the case of ties, the middle position is taken as the congruent
  registration position. Basis segments with a congruent registration
  position are called ``congruent matching profile segments'' (CMPS).
  The total number of CMPS is the CMPS score of the comparison.
  \texttt{get\_CMPS(input\_ccp,\ Tx\ =\ 25)} is the function that
  tallies the votes and determines the congruent registration position
  and congruent matching profile segments (CMPS).
\end{enumerate}

Note that there are several parameters in the CMPS algorithm that will
affect the final results and are left to the users to decide, such as
the length of basis segments \texttt{seg\_length} in step 1, the number
of peaks \texttt{npeaks\_set} identified on each level in step 2, and
the length of the tolerance zone \texttt{Tx} in both step 2 and 3. In
our implementation of the CMPS algorithm, we used the parameters given
in the original CMPS paper \citep{cmps} as the default values for these
parameters. However, the authors state that no cross-validation has been
done - there might also be issues with respect to the resolution of the
scans. Further research is needed, until then users are advised to think
of default values as starting values and consider alternatives. However,
the evaluation framework based on the Sum of Squares Ratio introduced in
the later section could be used to evaluate the choices of these
parameters. The main function that combines all steps in the CMPS
algorithm described above is called
\texttt{extract\_feature\_cmps(...)}. Here we present it with its
default parameters.

@@

The function \texttt{extract\_feature\_cmps} allows for the following
input from users besides the previously discussed parameters
\texttt{seg\_length}, \texttt{npeaks\_set}, and \texttt{Tx}:

\begin{itemize}
\item
  \texttt{x} and \texttt{y} are two signatures: \texttt{x} serves as the
  comparison signature (which will be divided into basis segments) and
  \texttt{y} is the reference signature;
\item
  \texttt{include} determines the format of the function result. Besides
  the \texttt{CMPS\_score}, other aspects of the comparison help in
  understanding how the \texttt{CMPS\_score} is computed. By default
  \texttt{include} is set to \texttt{NULL} and only the
  \texttt{CMPS\_score} is returned; further results are included when
  \texttt{include} is (an abbreviation of) one of or a vector of the
  following strings: \texttt{"nseg"}, \texttt{"congruent\_pos"},
  \texttt{"congruent\_seg\_idx"}, \texttt{"segments"},
  \texttt{"parameters"}, and \texttt{"full\_result"}. If
  \texttt{include} is specified as \texttt{"full\_result"} (or its
  abbreviation), the output includes everything listed below.

  \begin{itemize}
  \tightlist
  \item
    \texttt{nseg}: the number of basis segments from the comparison
    signature; this is also the highest possible CMPS score of the
    comparison;
  \item
    \texttt{congruent\_pos}: the congruent registration position;
  \item
    \texttt{congruent\_seg\_idx}: the indices of all congruent matching
    profile segments;
  \item
    \texttt{ccp\_list}: a list showing identified candidate positions of
    all basis segments;
  \item
    \texttt{pos\_df}: a data frame containing all candidate positions
    and their respective number of votes;
  \item
    \texttt{segments}: a list containing all basis segments;
  \item
    \texttt{parameters}: a list containing all input arguments of
    \texttt{extract\_feature\_cmps};
  \end{itemize}
\item
  \texttt{outlength} specifies the segment length of a basis segment at
  each level under the multi-segment lengths strategy. By default
  \texttt{outlength} is set to \texttt{NULL}, indicating that a basis
  segment should double its segment length at the next level and
  conforming to the description in \citet{cmps} .
\end{itemize}

In the remainder of the paper, we showcase the use of the \pkg{cmpsR}
functionality on some examples and present the results of applying it to
two datasets.
